Embedded Domain Specific Languages inside Functional Languages

### 2014-03-06 paper.tex
Adding text about Haskell.

>>>>

\section{Haskell Primer}

\section{Haskell}

Haskell is the premier pure functional programming language.
Haskell supports many forms of abstraction, and provides
a robust foundation for building Domain Specific Languages (DSLs).
In this section, we give a terse
introduction to Haskell,
sufficient to make this paper self-contained.

Haskell is all about {\em types\/}. Types in Haskell, like
types in other languages, are constraining summaries of structural values.
For example, in Haskell \verb|Bool| is the type of the values
\verb|True| and \verb|False|, \verb|Int| is the type of machine-sized
words, \verb|Double| is the type of double precision floating
point values; and this list goes on in the same manner as
\verb|C|, \verb|C++|, \verb|Java| and other traditional languages.
All these type names in Haskell start with an upper-case letter.

On top of these basic types, Haskell has two syntactical forms for expressing
compound types.
First, pairs, triples and larger structures can be written using tuple syntax,
comma separated types inside parenthesis.
So |(Int,Bool)| is structure with both an |Int| and a |Bool| component.
Second, lists have a syntactical shortcut, using square brackets.
So |[Int]| is a list of |Int|.

Haskell also has other container types. A container
that {\em may\/} contain one |Int| has the type
\verb|Maybe Int|
which is read |Maybe| of |Int|.
These container names also start with upper-case letters.
Types can be nested to any depth. For example, we can have a \verb|[(Maybe (Int,Bool))]|,
read as list of |Maybe| of (|Int| and |Bool|).

Polymorphic values, which are analogous to the type |Object| in Java,
or |void*| pointers in C, are expressed using lower-case letters.
These polymorphic values can have constraints expressed over them,
using the Haskell equivalent of an object hierarchy.
Finally, a Haskell function that takes a list, and returns a list
is written as using an arrow: |[a] -> [a]|.

We can now give an example of a Haskell function.%, the function \verb|sort|.
\begin{Code}

sort :: (Ord a) => [a] -> [a]
sort []     = []
sort (x:xs) = sort before ++ [x] ++ sort after
  where
        before = filter (<= x) xs
        after  = filter (> x) xs

\end{Code}
This function sorted a list, using a variant of quicksort in which the pivot is
the front of the list.
\begin{itemize}
\item The first line is the type for |sort|. This is $\forall$|a|, such that
|a| can be |Ord|ered (admits comparisons like |<=|), the function
takes and return a list of such |a|'s.
\item The second line says that an empty list is already sorted.
\item The remaining lines state that a (non-empty) list can be
sorted by taking the first and rest of the list (called |x| and |xs|, respectively),
and sorting the values before this pivot and after this pivot,
and concatenating theses intermediate values together.
\item Finally, intermediate values can be named using the |where| syntax;
in this case the values of |before| and |after|.
\end{itemize}

\pagebreak
Haskell is a concise and direct language.
Structures in Haskell are described using types, constructed
and deconstructed, but never updated. The entire language
functions by chaining together these structural processors,
which ultimately take input, and produce output. Side-effects
are described using a |do|-notation, for example:

main :: IO ()
main = do
  putStrLn "Hello"
  xs <- getLine
  print xs

\end{Code}

In this example a {\em value\/} called |main| uses the |do|-notation to describe
an interaction with a user. Actually, the |do|-notation captures
this as a structure called a Monad; purity is not compromised. For more details
about how |do|-notation and Monads can provide an effectful interface
inside a pure language like Haskell, see~\cite{SPJ:93:IFP}. For the
purposes of Kansas Lava, |do|-notation is a way of providing syntax
and structure that looks like interaction.
